From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-27 Pi+/paper/reversible1.tex
done with sec 3 for now

\noindent Intuitively, one may view each type $A$ as a collection of physical wires that can transmit $\sizet{A}$ distinct
values where $\sizet{A}$ is a natural number that indicates the size of a type, computed as: $\sizet{\zerot} = 0$;
$\sizet{\onet} = 1$; $\sizet{A + B} = \sizet{A} + \sizet{B}$; and $\sizet{A \times B} = \sizet{A} * \sizet{B}$.  Thus the type
$\mathbb{B} = \onet + \onet$ corresponds to a wire that can transmit one of two values, i.e., bits, with the
convention that $\inlv{\Acon{tt}}$ represents \AgdaFunction{𝔽} and $\inrv{\Acon{tt}}$ represents \AgdaFunction{𝕋}.  The
type $\mathbb{B} \times \mathbb{B} \times \mathbb{B}$ corresponds to a collection of wires that can transmit three
bits. From that perspective, a type isomorphism between types $A$ and $B$ (such that $\sizet{A}=\sizet{B}=n$) models a
\emph{reversible} combinational circuit that \emph{permutes} the $n$ different values. These type isomorphisms form a
\emph{commutative semiring} and are collected in Fig.~\ref{pi-terms}. Each line in the top part of the figure
introduces a pair of dual constants that witness the type isomorphism in the middle.  These are the \emph{base}
(non-reducible) terms of $\Pi$. The isomorphisms are extended to form a congruence relation by adding the three
constructors at the bottom of the figure that witness equivalence and compatible closure. Although austere, this
combinator-based language has the advantage of being amenable to formal analysis for at least two reasons: (i) it is
conceptually simple and small, and (ii) it has direct and evident connections to type theory and category theory and
hence possesses a rich algebraic structure which we exploit in the remainder of the paper.

@@

\paragraph*{Booleans}
Let us start with encoding booleans. We use the type \ensuremath{1{\sumtype}1} to
represent booleans with \ensuremath{\mathit{left} ~()} representing \ensuremath{\mathit{true}} and
\ensuremath{\mathit{right}~()} representing \ensuremath{\mathit{false}}.
Boolean negation is straightforward to define:

\ensuremath{\mathit{not} : \mathit{bool} \leftrightarrow \mathit{bool}}

\ensuremath{\mathit{not} = \swapp}

\noindent
It is easy to verify that \ensuremath{\mathit{not}} changes \ensuremath{\mathit{true}} to \ensuremath{\mathit{false}} and
vice versa.

\paragraph*{Bit Vectors.}
We can represent $n$-bit words using an n-ary product of
\ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
using the type \ensuremath{\mathit{bool} {\prodtype} (\mathit{bool} {\prodtype}  \mathit{bool})}.  We can perform various
operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
instance the bitwise \ensuremath{\mathit{not}} operation is the parallel composition of
three \ensuremath{\mathit{not}} operations:

\ensuremath{\mathit{not}_{\mathit{word}_3} :: \mathit{word}_3 \leftrightarrow \mathit{word}_3}

\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not}  {\prodtype}  (\mathit{not}  {\prodtype}  \mathit{not})}

\noindent We can express a 3-bit word reversal operation as follows:

\ensuremath{\mathit{reverse} : \mathit{word}_3 \leftrightarrow \mathit{word}_3}

\ensuremath{\mathit{reverse} = \swapt \odot (\swapt  \otimes  \idc)~ \odot \assocrt}

\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
many others we will see in this paper, is formed by sequentially
composing several simpler combinators. Instead of presenting the
operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
for presentation reasons) to flatten the tree into a sequence of
reductions as caused by each component. Such a sequence of reductions
is given below:
\[\begin{array}{rlr}
 & (v_1, (v_2, v_3)) \\
 \swapt & ((v_2, v_3), v_1) \\
 \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
 \assocrt & (v_3, (v_2, v_1)) \\
 \end{array}\]
%subcode source isomorphisms.tex:979

\noindent On the first line is the initial value. On each subsequent
line is a fragment of the \ensuremath{\mathit{reverse}} combinator and the value that
results from applying this combinator to the value on the previous
line. For example, \ensuremath{\swapt} transforms \ensuremath{(v_1, (v_2, v_3))} to
\ensuremath{((v_2,v_3),v_1)}.  On the last line we see the expected result with
the bits in reverse order.

We can also draw out the graphical representation of the 3-bit reverse
combinator. In the graphical representation, it is clear that the
combinator achieves the required shuffling.

% \inkscape{reverse-3-bit.pdf}

\paragraph*{Conditionals.}
Even though \ensuremath{\Pi } lacks conditional expressions, they are
expressible using the distributivity and factoring laws. The
diagrammatic representation of \ensuremath{\dist} shows that it redirects the flow
of a value \ensuremath{v:b} based on the value of another one of type
\ensuremath{b_1{\sumtype}b_2}. If we choose \ensuremath{1{\sumtype}1} to be
\ensuremath{\mathit{bool}} and apply either \ensuremath{c_1:b_1\leftrightarrow
b_2} or \ensuremath{c_2:b_1\leftrightarrow b_2} to the value \ensuremath{v},
then we essentially have an `if' expression.

\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool}  {\prodtype}  b_1 \leftrightarrow \mathit{bool}  {\prodtype}  b_2}

\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  \otimes\  c_1) {\sumtype} (\idc \otimes\  c_2)) \odot \factor}

@@

The diagram above shows the input value of type \ensuremath{(1{\sumtype}1) {\prodtype}  b_1}
processed by the distribute operator \ensuremath{\dist}, which converts it into
a value of type \ensuremath{(1 {\prodtype}  b_1){\sumtype}(1 {\prodtype}  b_1)}. In the
\ensuremath{\mathit{left}} branch, which corresponds to the
case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was
\ensuremath{\mathit{left} ~()}), the combinator~\ensuremath{c_1} is applied to
the value of type~\ensuremath{b_1}. The right
branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
the value of type \ensuremath{b_1} through the combinator \ensuremath{c_2}.
The inverse of \ensuremath{\dist}, namely \ensuremath{\factor} is applied
to get the final result of type \ensuremath{(1{\sumtype}1) {\prodtype} b_2}.

\paragraph*{Logic Gates}
There are several universal primitives for conventional (irreversible)
hardware circuits, such as \ensuremath{\mathit{nand}} and \ensuremath{\mathit{fanout}}. In the case
of reversible hardware circuits, the canonical universal primitive is
the Toffoli gate~\cite{Toffoli:1980}. The Toffoli gate takes three
boolean inputs: if the first two inputs are \ensuremath{\mathit{true}} then the third
bit is negated. In a traditional language, the Toffoli gate would be
most conveniently expressed as a conditional expression like:

\noindent
\ensuremath{ \mathit{toffoli}(v_1,v_2,v_3) = \mathit{if} ~(v_1 ~\mathit{and} ~v_2) ~\mathit{then} ~(v_1, v_2, \mathit{not}(v_3)) ~\mathit{else} ~(v_1, v_2, v_3)}

We will derive Toffoli gate in \ensuremath{\Pi } by first deriving a simpler
logic gate called \ensuremath{\mathit{cnot}}.  Consider a one-armed version, \ensuremath{\mathit{if}_c},
of the conditional derived above. If the \ensuremath{\mathit{bool}} is
\ensuremath{\mathit{true}}, the value of type \ensuremath{b} is modified by the operator \ensuremath{c}.

\begin{center}
\scalebox{1.5}{
%%subcode-line{pdfimage}[diagrams/if_c.pdf]
% \includegraphics{inkscape/cnot.pdf}
}
\end{center}

By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool} {\prodtype}  \mathit{bool}\leftrightarrow \mathit{bool} {\prodtype}  \mathit{bool}} which negates its
second argument if the first argument is \ensuremath{\mathit{true}}. This gate
\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.

If we iterate this construction once more, the resulting combinator
\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})\leftrightarrow \mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})}. The
resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},
proceeds to check the second argument. If that is also \ensuremath{\mathit{true}} then
it will negate the third argument. Thus \ensuremath{\mathit{if}_{\mathit{cnot}}} is the required
Toffoli gate.

\begin{center}
\scalebox{1.6}{
% \includegraphics{inkscape/toffoli.pdf}
}
\end{center}

>>>>

\noindent Intuitively, one may view each type $A$ as a collection of physical wires that can transmit $\sizet{A}$
distinct values where $\sizet{A}$ is a natural number that indicates the size of a type, computed as:
$\sizet{\zerot} = 0$; $\sizet{\onet} = 1$; $\sizet{A + B} = \sizet{A} + \sizet{B}$; and
$\sizet{A \times B} = \sizet{A} * \sizet{B}$.  Thus the type $\mathbb{2} = \onet + \onet$ corresponds to a wire that can
transmit one of two values, i.e., bits. The type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated
$\mathbb{B}~3$) corresponds to a collection of wires that can transmit three bits. From that perspective, a type
isomorphism between types $A$ and $B$ (such that $\sizet{A}=\sizet{B}=n$) models a \emph{reversible} combinational
circuit that \emph{permutes} the $n$ different values. These type isomorphisms form a \emph{commutative semiring} and
are collected in Fig.~\ref{pi-terms}. Each line in the top part of the figure introduces a pair of dual constants that
witness the type isomorphism in the middle.  These are the \emph{base} (non-reducible) terms of $\Pi$. The isomorphisms
are extended to form a congruence relation by adding the three constructors at the bottom of the figure that witness
equivalence and compatible closure. Although austere, this combinator-based language has the advantage of being amenable
to formal analysis for at least two reasons: (i) it is conceptually simple and small, and (ii) it has direct and evident
connections to type theory and category theory and hence possesses a rich algebraic structure which we exploit in the
remainder of the paper.

@@

In order to understand the examples of Sec.~\ref{sec:informal}, we review how to represent collections of booleans and
conditionals in $\Pi$. We use the type $\onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing
\textsf{false} and $\inrv{\Acon{tt}}$ representing $\textsf{true}$. Boolean negation is straightforward to define using
the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values.
For example, we can express a 3-bit word reversal operation as follows:

$\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$

$\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$

\noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
\[\begin{array}{rlr}
 & (v_1, (v_2, v_3)) \\
 \swapt & ((v_2, v_3), v_1) \\
 \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
 \assocrt & (v_3, (v_2, v_1)) \\
 \end{array}\]
%subcode source isomorphisms.tex:979

Even though \ensuremath{\Pi } lacks conditional expressions, they are expressible using the distributivity and factoring
laws as shown below:

$\Afun{ifc}_{c_1,c_2} : \mathbb{2} \times A \iso \mathbb{2} \times B$

$\Afun{ifc}_{c_1,c_2} = \dist \fatsemi ((\idc  \otimes\  c_1) \oplus (\idc \otimes\  c_2)) \fatsemi \factor$

@@

\noindent The input value of type $\mathbb{2} \times A$ is processed by the distribute operator \ensuremath{\dist},
which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the \ensuremath{\mathit{left}} branch,
which corresponds to the case when the boolean is \ensuremath{\mathit{false}}, the combinator~\ensuremath{c_1} is
applied to the value of type~\ensuremath{A}. The right branch which corresponds to the boolean being
\ensuremath{\mathit{true}} passes the value of type $A$ through the combinator \ensuremath{c_2}.  The inverse of
\ensuremath{\dist}, namely \ensuremath{\factor} is applied to get the final result. The full expressiveness of $\Pi$ is
demonstrated by the following implementation of $\Afun{cnot}$ and $\Afun{toffoli}$ which are universal for reversible
computation~\cite{Toffoli:1980}. The $\Afun{cnot}$ gates takes two input bits: a control bit and a target bit, and
negates the target bit if the control bit is set. It is essentially a one-branch version of $\Afun{cif}$ above. The
Toffoli gate generalizes the idea to a gate that takes 3 input bits: 2 control bits and one target bit: boolean inputs:
it negates the target bit if both control bits are set.

### 2021-06-27 Pi+/paper/reversible2.tex
rev1
done with sec 3 for now

\noindent Finally, given two combinators $p , q : A \isot_1 B$ and a 2-combinator $\alpha : p \iso_2 q$, the denotation
$\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.

>>>>

\noindent Finally, given two combinators $p , q : A \isot_1 B$ and a 2-combinator $\alpha : p \Iso q$, the denotation
$\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.
